---
title: "lubridate_fill_in"
author: "Stephanie Copeland"
date: "2024-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### Packages ####
library(here)
library(tidyverse)
library(ggplot2)
library(lubridate)

#### Read in initial raw data .csv ####
mos_counts_raw <- read_csv(
  here::here("mos_counts_2023.csv"))
```

### Raw Data Information ###
- Mosquito collections taken around 10 alpine lakes in Inyo National Forest from July 15th till August 25th.

#### Here are the columns and their associated characteristics ####
- sampler - character: Initial ID of which researcher was sampling
- SC = Steph Copeland (project lead)
- MR = Matthew Rosen (undergraduate techician)
- SL = Sylvia Li (undergraduate technician)
- ST = Sachi Takishima (volunteer)
- CC = Charlie Copeland (volunteer, Steph's dad)
- KK = Kayla Kauffman (volunteer)
- *also present for all sampling endeavors but not a sampler was Hugo, Steph's service dog. He is a 70lbs German Shepherd Mix*
- Lake - character: name of lake which collections were taken. *Heart, Hidden, Flower, No_Name, Gilbert, Matlock, Marsh, LEB (Lower Eastern Brook), Little_Fish, Little_Lakes, Mack, EB (Eastern Brook)*
- Date - character: changed to date class (see below) date at which sampling occurred
- Time_start - time: time that sampling began
- Time_end = time: time that sampling commenced
- Total_time - numeric: the amount of minutes sampled, calculated by determining the number of minutes that had passed between Time_start and Time_end
- snow_melt_mosq - numeric: count of caught mosquitos falling into this *phenotypic category*
- black_mosq - numeric: count of caught mosquitos falling into this *phenotypic category*
- culiseta - numneric: count of mosquitos falling into this *genera* 
- culex_tarsalis - numeric: count of caught mosquitos falling into this *species* 
- blood_test - numeric: number of mosquitos removed from each sampling effort to test for the content of a blood meal. These samples were inherently destroyed by puncturing the abdomen of the sample with a burn/sterile pipette tip 

#### clean names to make column all lower case ####
- this is optional, but nice so you don't have to think about capitalization while you're coding
```{r}
mos_counts_clean <- mos_counts_raw %>%
  janitor::clean_names()
```

#### date cleaning and transforming with 'lubridate' ####
- as you can see, the date column does not include year
- also, single digit dates e.g. August 3rd are represented by 3-Aug, not 03-Aug
- it is also currently recognized by R as a 'chr' column, not a 'date' colum
- these are the things we need to change!
```{r}
### add a zero infront of any "single digit" dates with 'gsub' ###
mos_counts_clean$date <- gsub("(\\b\\d{1}\\b)", "0\\1", mos_counts_clean$date)

### remove dash b/w day and month ###
mos_counts_clean$date <- gsub("-", "", mos_counts_clean$date)

### change month to number and not letter code ####
mos_counts_clean$date <- gsub("Jul", "07", mos_counts_clean$date)
mos_counts_clean$date <- gsub("Aug", "08", mos_counts_clean$date)

### add 2023 to each cell in the date column ####
mos_counts_clean$date <- paste(mos_counts_clean$date, "2023", sep = "")
```

- 'gsub' is a base R function - *tbh I use chatGPT a lot when using it to get the syntax right*
- now the data is in a format day, month, year (dmy) that 'lubridate' can recognize and can transform this column into a 'date' column, not a 'chr' column. 

#### transforming date column ####\
```{r, fill in, lubridate practice!}

```
- 'lubridate' has a lot more functions than just this one.
- A cheat sheet on 'lubridate' should be in this weeks repro for you!

### with dates you can do time-based visualizations or analysis ###

#### but first, more data cleaning to do so ####
##### More background on this work #####
- These lakes were put in "pairs" as lakes containing fish and lakes containing no fish. 
- 90% of the Eastern Sierra Nevada have had high alpine lakes artificially stocked with non-native trout starting in the late 1800s. 
- These trout are voracious predators and have caused a litany of top-down as well as cross-ecosystem impacts.
- One area previously not researched in great detail, is their impact to mosquito populations
```{r}
### add column that designates lakes a "fish" or "no fish"
mos_counts_clean <- mos_counts_clean %>% 
  mutate(fish_status = case_when(
    lake %in% c("EB", "Hidden", "LEB", "Little_Lakes", "Matlock", "No_Name") ~ "no_fish",
    TRUE ~ "fish"
  ))

#### group by lake and date ####
mos_counts_time_vis <- mos_counts_clean %>% 
  group_by(lake, date) %>% 
  summarise(
    time_min = sum(total_time),
    total_snow_melt_mosq = sum(snow_melt_mosq),
    total_black_mosq = sum(black_mosq),
    total_culiseta = sum(culiseta, na.rm = TRUE),
    total_culex_tarsalis = sum(culex_tarsalis),
    total_blood_test = sum(blood_test, na.rm = TRUE)
  )

### add column of total mosquitoes by lake and date ###
mos_counts_time_vis <- mos_counts_time_vis %>% 
  mutate(total_sum = total_snow_melt_mosq + 
           total_black_mosq +
           total_culiseta +
           total_culex_tarsalis +
           total_blood_test)

### create a column of mosquito captures per collection minute, this is a rate and can be used in analyses when sampling is uneven or not continuous ###
mos_counts_time_vis <- mos_counts_time_vis %>% 
  mutate(effort_min = total_sum/time_min)

### round decimal place to 1 ###
mos_counts_time_vis$effort_min <- round(mos_counts_time_vis$effort_min, 1)
```

#### now to visualize! ####
```{r}

```


```{r}

```